---
title: "할수 있는거 예시"
author: ""
date: "2025-05-14"
output:
  pdf_document:
    toc: true          
    number_sections: true  
    fig_caption: true  
    df_print: kable    
    latex_engine: xelatex  
mainfont: "맑은 고딕"   
fontsize: 11pt
geometry: margin=1in
---
```{r setup, include = FALSE}
library(tidyverse)
library(modelr)
library(boot)
library(infer)
library(HistData)
```


```{r}
student_habits_performance <- read.csv("student_habits_performance.csv")
```



#### 1. Visualization data

### example code

### scatter plot
### Relationship between Study Time and Exam Score


```{r}
student_habits_performance %>% ggplot() +
  geom_point(aes(x = study_hours_per_day, y = exam_score))
```



### Histogram
### Distribution of Sleep Time

```{r}
student_habits_performance %>% ggplot() +
  geom_histogram(aes(x = sleep_hours), bins = 30)
```






#### 2. Data Wrangling

- Creating new variables by using mutate



#### example code

```{r}
student_habits_performance_new <- student_habits_performance %>% mutate(study_sleep_ratio = study_hours_per_day / sleep_hours)
```


#### 3. Data Filtering

### example code

```{r}
head(student_habits_performance %>% filter(sleep_hours >= 8, exam_score >= 90)) %>% 
  select(age, study_hours_per_day, sleep_hours, exam_score)
```


#### 4. inference

#### Example

- Is there an average grade difference between students with long study time
and those with short study time?

The null hypothesis (H0): There is no difference in average grade between the group with a lot of study time and the group with a little study time.
The alternative hypothesis (H1): There is a difference in average grade between the two groups.



#### 5. Linear Regression

#### example

공부시간이 많을수록 성적이 오를까??




